---
title: "Test crop PNA Illumina Run Dec2022"
author: "Briana K. Whitaker"
date: "`r Sys.Date()`"
output: html_document
---
\fontsize{9}{10}
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3.5, fig.height=3,
                      warning=FALSE, message=FALSE)
```


* Run using `r version[['version.string']] `.


# Objective
This document reports on results from the Illumina Nano Run in Dec 2022, on 40 samples. 1 primer pair only. 3 wheat head, 3 barley head, and 3 corn leaf samples times four PNA treatment levels Plus 2 corn debris samples as positive controls [no plant amplification in previous trials], and two PCR negative controls.


# Load Packages, set path to data
```{r, echo=FALSE, results='hide', include=FALSE} 

x<-c("Biostrings", "car", "genefilter", "MASS",
    "tidyverse", "ggplot2", "phyloseq", "vegan",
    "betareg", "brms", "bayesplot", "posterior") # these 4 needed for beta-reg models only
lapply(x, require, character.only = TRUE)

# require(devtools); packageVersion("devtools")               #v.‘2.4.5’
# devtools::install_github("shenhav-and-korem-labs/SCRuB")
require(torch); packageVersion("torch")                     #v.‘0.15.0’
require(SCRuB); packageVersion("SCRuB")                     #v.‘0.0.1’


#Load functions
source("./code/fxn_blastr-bw.R")
source("./code/blastn_code.R")
source("./code/multiplot.function.R")

#add 'not in' function
`%nin%` = Negate(`%in%`)

#set seed
set.seed(217)

# set ggplot2 theme
theme_set(theme_bw(base_size=10)) 
theme_update(panel.grid.major=element_line(0), panel.grid.minor=element_line(0))
```


```{r, echo=FALSE, results='hide', include=FALSE} 

SbyE <- read.csv("./data/cropPNA_TestIllumina_2022-12-02.csv")

track <- read.csv("./data/cropPNA_TrackSequences.csv", 
                 row.names = 1)
SbyS <- read.csv("./data/cropPNA_SbyS.csv", row.names=1)

uniqueSS <- readDNAStringSet("./data/cropPNA_uniqueSeqs.fasta")

rownames(SbyE) <- SbyE$trtID
SbyE$trtID <- as.factor(SbyE$trtID)
SbyE$ID <- as.factor(SbyE$ID)
SbyE$Crop.Type <- as.factor(SbyE$Crop.Type)
SbyE$PNA_Treatment_Code <- as.factor(SbyE$PNA_Treatment_Code)
SbyE$PNA_Treatment <- as.factor(SbyE$PNA_Treatment)
SbyE$Replicate <- as.factor(SbyE$Replicate)
SbyE$Index_N7XX <- as.factor(SbyE$Index_N7XX)
SbyE$Index_S5XX <- as.factor(SbyE$Index_S5XX)
# drop PCRneg 2, as no reads passed filter
SbyE <- SbyE[SbyE$trtID != "PCRneg-2",]

str(SbyE)

# sanity check, must be true
identical(sort(rownames(SbyE)), sort(rownames(SbyS)))


load("./data/cropPNA_ps_ids.RData")
ps

conf <- read.csv("./data/cropPNA_decipher_ids_conf.csv", row.names = 1)


# add track sequence data to phyloseq obj
identical(sort(rownames(track)), sort(sample_names(ps))) #must be TRUE

track_ps <- phyloseq(sample_data(track))

ps <- merge_phyloseq(ps, track_ps)
ps

```



# Summarize & Remove Plant Reads

### Run hits from DECIPHER with low confidence at Kingdom level on ncbi browser
```{r, results='hide', echo=FALSE}
lowConfASVs <- rownames(conf)[conf$k.conf<70]

# lowConfSS <- uniqueSS[names(uniqueSS) %in% lowConfASVs]
# writeXStringSet(lowConfSS, "./data/cropPNA_uniqueSeqs_lowConfDecipher.fasta")
lowConfSS <- readDNAStringSet("./data/cropPNA_uniqueSeqs_lowConfDecipher.fasta")
hist(width(lowConfSS)) 
range(width(lowConfSS)) #couple short ones too
```

* ASVs that are low confidence from decipher output are generally longer than average for regular dataset ~50bp

```{r, results='hide'}
# ##write.csv(lowConfASVs, "./data/cropPNA_lowConfASVs.csv") #manually edited with blast-nr results
lowConf.ncbi <- read.csv("./data/cropPNA_lowConfASVs.csv", row.names = 1, stringsAsFactors = TRUE)

# table(lowConf.ncbi$Problematic)      # 16 problematic overall
# table(lowConf.ncbi$NCBI.Blast.Group) # 10 plant (leaving 6 as chimera seqs)
# table(lowConf.ncbi$NCBI.Blast.ID)    #3 Hordeum, 6 Triticum, 1 Hord/Trit chimera
```

* Overall 16 residual ASVs that are likely plant or chimera. Rest are Fungi, just weren't determined from DECIPHER matches.

```{r, results='hide', echo=FALSE}
notFungiASVs <- droplevels(lowConf.ncbi$ASV[lowConf.ncbi$Problematic=="yes"])
plantASVs <- droplevels(lowConf.ncbi$ASV[lowConf.ncbi$NCBI.Blast.Group=="Plant"])

test <- ps@otu_table@.Data

propPlant <- rowSums(test[,colnames(test) %in% plantASVs])
ps@sam_data$propPlant <- c(propPlant/rowSums(ps@otu_table))

```

* 3 ASVs were Hordeum, while 6 were Triticum, and 1 was a Triticum-Hordeum chimera of sorts. In addition, 6 ASVs were likely chimeras


### subset phyloseq obj
```{r, echo=FALSE}
ps_fungi <- ps
ps_fungi@otu_table <- otu_table(ps)[,colnames(otu_table(ps)) %nin% notFungiASVs]
ps_fungi@refseq <- refseq(ps_fungi)[names(refseq(ps_fungi)) %nin% notFungiASVs]
#ps_fungi  #removes the 16 problematic ASVs

# add new seq depth
ps_fungi@sam_data$seqDepth_noPlant <- rowSums(ps_fungi@otu_table)

#save(ps_fungi, file = "./data/cropPNA_ps_fungi.RData")
load("./data/cropPNA_ps_fungi.RData")
ps_fungi
```

```{r}
sum(ps_fungi@otu_table['PCRneg-1',])
#sum(ps@otu_table['PCRneg-1',])

```

* Note, 243 reads in PCRneg1 sample (no reads passed filter for PCRneg2)
* None of the reads in the PCRneg1 are plant-origin. 


### Plots of sequencing and plant contamination

```{r, include=FALSE}
data.frame(ps_fungi@sam_data@.Data) -> test0
colnames(test0) <- ps_fungi@sam_data@names
rownames(test0) <- ps_fungi@sam_data@row.names
str(test0)

test <- test0 %>% filter(Crop.Type != 'Debris') %>% filter(trtID != 'PCRneg-1')
test$PNA_Treatment <- factor(test$PNA_Treatment, levels = c("0uM", "0.75uM", "1uM", "1.5uM"))

test %>% droplevels() -> test

```

##### Proportion of sample that is plant reads
```{r, results='hide', echo=FALSE, fig.width = 6, fig.height=4}
p1 <- ggplot(test, aes(y = propPlant, x = PNA_Treatment, color = Crop.Type)) +
  facet_grid(.~Crop.Type) + 
  geom_violin() + geom_jitter(size = 1.5, width = 0.2) +
  guides(color = 'none') +
  scale_x_discrete("PNA Treatment") + 
  scale_y_continuous("Proportion that are Plant Reads", trans = "sqrt") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
#tiff("./figures/Fig2_cropPNA.tiff", width=3.25, height=3.25, units="in", res=300)
p1
#dev.off()
```

##### Raw Sequencing depth for sample after DADA2, prior plant removal
```{r, results='hide', echo=FALSE, fig.width = 6, fig.height=4}
ggplot(test, aes(y = postChimera , x = Crop.Type)) +
  ylab("raw Seq. Depth") +
  geom_boxplot() + geom_jitter(size = 3, width = 0.1, aes(color=PNA_Treatment))

ggplot(test, aes(y = postChimera, x = PNA_Treatment, color = Crop.Type)) +
  facet_grid(.~Crop.Type)+ 
  ylab("raw Seq. Depth") +
  geom_violin() + geom_jitter(size = 3, width = 0.1)

```

##### Sequencing depth for sample after plant removal
```{r, results='hide', echo=FALSE, fig.width = 6, fig.height=4}
ggplot(test, aes(y = seqDepth_noPlant , x = Crop.Type)) +
  geom_boxplot() + 
  geom_jitter(size = 3, width = 0.1, aes(color=PNA_Treatment)) + 
  ylab("(no Plant) Seq. Depth")
ggplot(test, aes(y = seqDepth_noPlant, x = PNA_Treatment, color = Crop.Type)) +
  facet_grid(.~Crop.Type)+ 
  ylab("(no Plant) Seq. Depth") +
  geom_violin() + 
  geom_jitter(size = 3, width = 0.1)
```


* biggest effect from crop type, 


### Run Models
```{r, echo = FALSE}
# create a continuous variable
test$PNA_Treatment_cont <- as.numeric(
  ifelse(test$PNA_Treatment == "0uM", "0", 
  ifelse(test$PNA_Treatment == "0.75uM", "0.75",
  ifelse(test$PNA_Treatment == "1uM", "1",
  ifelse(test$PNA_Treatment == "1.5uM", "1.5", NA)))) )
# remove 0 boundary observations from proportion plant reads variable
test$propPlant_01 <- test$propPlant+0.01


# general linear model
plant.mod1 <- lm(propPlant ~ Crop.Type*PNA_Treatment, data = test)
car::Anova(plant.mod1, type = 2)
#par(mfrow = c(2,2)); plot(plant.mod1); par(mfrow = c(1,1))
# poor model fit

# Beta regression model for proportions
### note, regular betareg can not handle boundary observations [0,1], so extended distribution is used dist='xbetax', which has issues with testing the interaction. Added small proportion 0.01 to allow test of interaction
plant.mod2 <- betareg(I(propPlant+0.01) ~ Crop.Type*PNA_Treatment, data = test,  
        phi = FALSE, link = "logit", na.action = na.omit, type = "ML",
        dist = 'beta')
car::Anova(plant.mod2, type = 2)
#par(mfrow = c(2,2)); plot(plant.mod2); par(mfrow = c(1,1))
# poor model fit

# plant.mod3 <- betareg(propPlant ~ Crop.Type*PNA_Treatment, data = test,  
#         phi = FALSE, link = "logit", na.action = na.omit, type = "ML",
#         dist = 'xbeta', nu = 0.0002)
# #car::Anova(plant.mod3, type = 2)
# par(mfrow = c(2,2)); plot(plant.mod3); par(mfrow = c(1,1))


plant.mod4 <- betareg(propPlant ~ Crop.Type+PNA_Treatment_cont, data = test,  
        phi = FALSE, link = "logit", na.action = na.omit, type = "ML",
        dist = 'xbetax')
car::Anova(plant.mod4, type = 2)
#par(mfrow = c(2,2)); plot(plant.mod4); par(mfrow = c(1,1)) 
# good fit, but no interaction tested

plant.mod5 <- betareg(I(propPlant+0.01) ~ Crop.Type*PNA_Treatment_cont, data = test,  
        phi = FALSE, link = "logit", na.action = na.omit, type = "ML",
        dist = 'beta')
car::Anova(plant.mod5, type = 2)
#par(mfrow = c(2,2)); plot(plant.mod5); par(mfrow = c(1,1))
# poor model fit

# plant.mod6 <- betareg(I(propPlant+0.01) ~ Crop.Type*PNA_Treatment_cont, 
#         data = test[test$Crop.Type != "Corn",],  
#         phi = FALSE, link = "logit", na.action = na.omit, type = "ML",
#         dist = 'beta')
# car::Anova(plant.mod6, type = 2)
# par(mfrow = c(2,2)); plot(plant.mod6); par(mfrow = c(1,1))
# poor model fit

# quasibinomial model (sometimes for proportions) ### https://www.andrewheiss.com/blog/2021/11/08/beta-regression-guide/)
plant.mod7 <- glm(propPlant ~ Crop.Type*PNA_Treatment, data = test,
                  family = quasibinomial())
car::Anova(plant.mod7, type = 2)
#par(mfrow = c(2,2)); plot(plant.mod7); par(mfrow = c(1,1))  
# poor model fit
```


```{r}
## Bayesion ZOIB (zero inflated beta regression)
bf.mod <- bf(propPlant ~ Crop.Type*PNA_Treatment_cont,
     zi ~ Crop.Type) #phi ~ Crop.Type*PNA_Treatment_cont)
get_prior(
  bf.mod,
  data = test,
  family = zero_inflated_beta()
)
# don't want all the 'flat' priors
my.priors <- c(set_prior("student_t(3, 0, 2.5)", class = "Intercept"),
                    #keep the default prior for the intercept
               set_prior("normal(0, 2)", class = "b"))
                    #but add priors for the b class
plant.mod9 <- brm(bf.mod,
  data = test,
  family = zero_inflated_beta(),
  init = 0,
  #chains = 4, iter = 20, warmup = 10,
  #cores = 4, seed = 217, 
  chains = 4, iter = 2000, warmup = 1000,
  cores = 4, seed = 217, 
  backend = "rstan",
  prior = my.priors,
  file = "./data/plant.mod9"  
)
plant.mod9

## Model convergence
# inspect trace plots (for MCMC convergence)
mcmc_trace(plant.mod9)  #fuzzy caterpillae, all chains mixed and yielded the same estimate + approx credible interval
# inspect Rhats
plant.mod9.rhats <- rhat(plant.mod9)
mcmc_rhat(plant.mod9.rhats)  # want rhats to be as close to 1 as possible
# compares variance w/i chains to vairance b/w chains
# values below 1.01 are good, which all of these are < 1.005
table(plant.mod9.rhats<1.01)
# also there were no warnings about the bulk or Tail effective sample size (ESS), which are all above at least 1000 in this model
# and no warnings about transitions exceeding the max. treedepth

# examine estimates
plogis(-1.35)
# 20% increase in proportion plant reads in barley
plogis(-1.35 + -1.80) - plogis(-1.35) 
#16% decrease in proportion plant reads in wheat relative to barley
# note - corn B estimate includes 0 in CI, so probably most of the information for corn is contained in the zi parm
plogis(-1.35 + -2.87) - plogis(-1.35) 
# 19% decrease in proportion plant reads with increase in PNA concentration of 1uM. Note - this is only true for default intercept which is barley, both the interaction terms with corn and wheat include 0 in the credible intervals

plogis(-2.00)
#12% 0s in barley (which is about accurate)
plogis(-2.00 + 8.45) - plogis(-2.00) 
#88% increase in 0s in corn relative to barley
plogis(-2.00 + 3.11) - plogis(-2.00) 
#63% increase in 0s in corn relative to barley


# # try to include interaction
# bf.mod2 <- bf(propPlant ~ Crop.Type*PNA_Treatment_cont,
#               zi ~ Crop.Type*PNA_Treatment_cont) #phi ~ Crop.Type*PNA_Treatment_cont)
# plant.mod10 <- brm(bf.mod2,
#   data = test,
#   family = zero_inflated_beta(),
#   init = 0,
#   #chains = 4, iter = 20, warmup = 10,
#   #cores = 4, seed = 217, 
#   chains = 4, iter = 2000, warmup = 1000,
#   cores = 4, seed = 217, 
#   backend = "rstan",
#   prior = my.priors,
#   file = "./data/plant.mod10"  
# )
# plant.mod10
# # lots of warnings about low ESS, large Rhats, and transitions exceeding max treedepth
# #mcmc_trace(plant.mod10) #looks really bad
# # poor model convergence

```

* Both crop type and PNA treatment have an effect on the proportion of plant reads contaminating a sample. 
*  Normal linear model implies: Barley 0uM worse than any other barley trt (L,M,H), and also barley 0uM worse than all corn and wheat trts (all 0,L,M,H). But, though the wheat 0uM looks higher, it is not statistically different from wheat PNA trts (L,M,H)
* Bayesian regression model largely supports this lm model interpretation

```{r}
raw.mod1 <- lm(postChimera ~ Crop.Type*PNA_Treatment, data = test)
car::Anova(raw.mod1, type = 2)  # only crop type has an effect on seq depth, not inclusion of the PNA

depth.mod1 <- lm(seqDepth_noPlant ~ Crop.Type*PNA_Treatment, data = test)
car::Anova(depth.mod1, type = 2)
#depth.aov <- aov(seqDepth_noPlant ~ Crop.Type*PNA_Treatment, data = test)
#TukeyHSD(depth.aov, which = "Crop.Type")  #corn diff from barley, but not wheat, wheat not diff from barley

```

* However, only crop type has an effect on the total number of raw reads that pass through the DADA2 pipeline, and also the only significant effect on the total sequence depth after removal of plant reads.


# Examine Controls
```{r}
ps_cont <- subset_samples(ps_fungi, trtID == "PCRneg-1")
ps_cont <- filter_taxa(ps_cont, function(x) mean(x) > 0, TRUE)
otu_table(ps_cont)

bad_taxa <- taxa_names(ps_cont)

```

* 9 ASVs in controls. More than usual.

```{r}
ps_fungi@tax_table[bad_taxa] # 14 reads
# variable taxonomy
#sort(rowSums(ps_fungi@otu_table[,bad_taxa]), dec = TRUE)
## reads that appear in the negative, appear in every other sample run.
```

* Small problem with contamination in this run. Use SCRuB to decontaminate.

```{r}
meta <- (as(sample_data(ps_fungi), "data.frame"))
mat <- (as(otu_table(ps_fungi), "matrix"))

# format for SCRuB
meta$sample_well <- as.character(meta$Well)
meta$is_control <- ifelse(meta$trtID== "PCRneg-1", "TRUE", "FALSE")
meta$sample_type <- ifelse(meta$trtID== "PCRneg-1", "NegControl", 
                           ifelse(meta$Crop.Type== "Debris", "Debris", "Phyllosphere"))
meta %>% arrange(trtID) %>% select(is_control, sample_type, sample_well) -> meta
mat %>% data.frame %>% mutate(ids = rownames(mat)) %>%
    arrange(ids) %>% select(-ids) %>% as.matrix -> mat
#identical(rownames(mat), rownames(meta)) #TRUE

# scr_out <- SCRuB(mat, meta, control_order = "NegControl" )  #with spatial info aka well info
# save(scr_out, file="./data/cropPNA_scrubOut.RData")
load("./data/cropPNA_scrubOut.RData")

# Estimated contamination level #read boxplot as contam proportion
(1 - scr_out$p) %>% boxplot()              
median(100-(scr_out$p*100))  #0%  (in percentages)
# Estimated realtive abundance of the contamination community
scr_out$inner_iterations$`NegControl`$gamma %>% plot()  # 4 asvs w/ some higher contam
### Estimate well-to-well leakage from PCR controls
scr_out$inner_iterations$`NegControl`$alpha[, ncol(scr_out$inner_iterations$`NegControl`$alpha)]
# 0.05% well-to-well leakage across plate

# final SbyS mat cleaned up
decon <- scr_out$decontaminated_samples %>% as.matrix()
dim(decon) # removes neg. control, 38 samples remain, 264 taxa (loss of 16 taxa)
#dim(SbyS)
```


# Diversity
* For the statistical models, I only compared the wheat head, barley head, and corn leaf samples, not the debris or negative control. 

### H Diversity
```{r, fig.width = 8, fig.height = 4}
# format for diversity
SbyE <- SbyE[order(SbyE$trtID),] %>% filter(trtID != "PCRneg-1")
#identical(rownames(SbyE), rownames(decon)) #must be TRUE
SbyE$Diversity <- diversity(decon, index = 'shannon')
SbyE2 <- SbyE %>% filter(Crop.Type != 'Debris')
SbyE2$PNA_Treatment <- factor(SbyE2$PNA_Treatment, 
                              levels = c("0uM", "0.75uM", "1uM", "1.5uM"))

p2 <- ggplot(SbyE2, aes(x = PNA_Treatment, y = Diversity, color = Crop.Type)) +
  facet_grid(.~Crop.Type) +
  geom_violin() + geom_jitter(size = 1.5, width = 0.2) +
  guides(color = 'none') +
  scale_x_discrete("PNA Treatment") + 
  scale_y_continuous("Shannon Diversity", trans = "sqrt") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#tiff("./figures/Fig3a_cropPNA.tiff", width=3.25, height=3.25, units="in", res=300)
p2
#dev.off()

h.mod1 <- lm(Diversity ~ Crop.Type*PNA_Treatment, data = SbyE2)
#par(mfrow=c(2,2))
#plot(h.mod1)
car::Anova(h.mod1, type = 2)
```



### Structure
```{r, fig.width = 7, fig.height = 5}
SbyS2 <- decon[rownames(decon) %in% rownames(SbyE2),]
SbyS2 <- SbyS2[,colSums(SbyS2) >0]    #29 ASVs unique to debris
#dim(decon); dim(SbyS2)
SbyS2.hel <- decostand(SbyS2, method = 'hellinger')
dist <- vegdist(SbyS2.hel, method = 'bray')
pcoa <- cmdscale(dist, eig =TRUE)
expl_1 <- round(pcoa$eig[1] / sum(pcoa$eig), 3) * 100
expl_2 <- round(pcoa$eig[2] / sum(pcoa$eig), 3) * 100 
expl_3 <- round(pcoa$eig[3] / sum(pcoa$eig), 3) * 100 
scores <- as.data.frame(pcoa$points)
scores$ids <- rownames(scores) 
pdat <- merge(scores, SbyE2, by.x="ids", by.y="trtID")
rownames(pdat) <- as.character(pdat$ids)

p3 <- ggplot() + coord_equal() + 
    geom_point(data = pdat, aes(x = V1, y = V2, 
          color = Crop.Type, shape = PNA_Treatment) , size = 1.5) +
    scale_x_continuous(paste("PCoA 1 (", expl_1, "%)", sep = "")) +
    scale_y_continuous(paste("PCoA 2 (", expl_2, "%)", sep = "")) +
  scale_color_discrete("") + scale_shape_discrete("") +
  guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2)) +
  theme(legend.position = 'top', 
        legend.spacing.x = unit(0.0, "cm"),
        legend.margin = margin(0, 0, 0, 0, unit = "cm"))
#tiff("./figures/Fig3b_cropPNA.tiff", width=3.25, height=3.25, units="in", res=300)
p3
#dev.off()

#test interaction
out <-adonis2(SbyS2.hel ~ Crop.Type*PNA_Treatment, data = SbyE2, 
              method = 'bray', by = "margin")  
out   #not sgnif
# test main effects
out2 <-adonis2(SbyS2.hel ~ Crop.Type+PNA_Treatment, data = SbyE2, 
             method = 'bray', by = "margin")  
out2 

# NOTE - vegan::adonis2 has changed versions/formulas since this script was first written; so model script has changed, but results are the same.
```

*Only crop type affects the community structure and shannon diversity, not PNA test (=good!)

```{r, include = FALSE}
#tiff("./figures/Fig3_cropPNA.tiff", width=7, height=3.5, units="in", res=300)
multiplot(p2+ggtitle("A"), p3+ggtitle("B"), cols = 2)
#dev.off()

```


```{r}
sessionInfo()
```


### end